---
title: "Live and Large Plant Phytochemistry Queries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Live and Large Plant Phytochemistry Queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Scope

This article documents optional live-provider work. Its code is not executed
during package checks or website builds. Public services change, throttle
requests, and occasionally become unavailable. Run a bounded pilot, retain
provider diagnostics, and reuse persistent caches before scaling.

Public-source results are reported evidence, not a complete metabolome and not
proof of chemistry in a new experimental sample.

## Configure NCBI responsibly

Set NCBI contact information in the R session or operating-system environment.
The API key is optional. Never commit it or place it in exported manifests.

```{r ncbi-config, eval=FALSE}
Sys.setenv(
  NCBI_EMAIL = "researcher@example.edu",
  NCBI_TOOL = "uafR"
)

# Optional:
# Sys.setenv(NCBI_API_KEY = "set-this-outside-version-control")
```

Without an API key, keep NCBI requests at or below three per second. uafR uses
conservative defaults, caching, backoff, and resumable service-busy handling,
but the user remains responsible for appropriate API use.

## Start with a three-species pilot

```{r small-live-query, eval=FALSE}
library(uafR)

plants = c("Salix nigra", "Camellia sinensis", "Zea mays")

phyto = resolvePlantPhytochemistry(
  plants = plants,
  sources = c("lotus", "pubmed", "pubtator"),
  taxon_fallback = c("species", "genus"),
  enrich_compounds = FALSE,
  cache = TRUE,
  cache_dir = "uafR_plant_cache",
  throttle = 0.5,
  max_pubmed_records = 25,
  max_provider_records = 100,
  progress = TRUE
)

phyto$ProviderDiagnostics
phyto$PlantCompoundOccurrences
phyto$LiteratureCandidates
validatePlantPhytochemistryResult(phyto)$Summary
```

Confirm that each enabled provider is reported as queried, unavailable,
disabled, no-hit, incomplete, or successful. A transport error must not be
interpreted as a biological no-hit.

## Prefer a local LOTUS index

For medium or large panels, download an official LOTUS export once, preserve its
source URL and checksum, and build a local index:

```{r local-lotus, eval=FALSE}
lotus_build = buildLotusIndex(
  input = "downloads/lotus_flat_export",
  out_file = "uafR_indexes/lotus_index.rds",
  overwrite = FALSE
)

lotus_hits = queryLotusIndex(
  plants = plants,
  lotus_index = "uafR_indexes/lotus_index.rds",
  taxon_fallback = c("species", "genus"),
  max_records = 500
)

phyto = resolvePlantPhytochemistry(
  plants = plants,
  sources = c("lotus", "pubmed", "pubtator"),
  lotus_index = "uafR_indexes/lotus_index.rds",
  enrich_compounds = FALSE,
  cache = TRUE,
  cache_dir = "uafR_plant_cache"
)
```

Use `tools/flatten_lotus_mongo_dump.py` for official MongoDB ZIP exports when a
flat file is not already available.

## Preflight a large panel

`planPlantChemistryRun()` validates input names and estimates request and
pairwise-output burden without making live requests:

```{r large-preflight, eval=FALSE}
run_plan = planPlantChemistryRun(
  plants = "project_species.csv",
  sources = c(
    "lotus",
    "npass",
    "knapsack",
    "pubchem",
    "pubmed",
    "pubtator"
  ),
  cache_dir = "uafR_plant_cache",
  lotus_index = "uafR_indexes/LOTUS_lookup_index",
  expected_compounds_per_plant = 25,
  write_full_pairwise = FALSE,
  species_chunk_size = 25,
  compound_batch_size = 25,
  max_pubmed_records = 25
)

run_plan$InputNameAudit
run_plan$ProviderPlan
run_plan$CacheSummary
run_plan$OutputEstimates
run_plan$ReadinessChecks
run_plan$Recommendations
```

Resolve blank, duplicate, genus-only, and `sp.`/`spp.` names before discovery.
Supply reviewed taxonomy metadata rather than inferring family or accepted
names without a source.

## Use the resumable production panel

Run preflight and pilot as separate gates:

```sh
Rscript tools/run_plant_chemistry_panel.R \
  --mode preflight \
  --plant-csv project_species.csv \
  --out-dir plant_phytochemistry_panel \
  --cache-dir uafR_plant_cache \
  --lotus-index uafR_indexes/LOTUS_lookup_index \
  --sources lotus,npass,knapsack,pubchem,pubmed,pubtator \
  --species-chunk-size 25 \
  --compound-batch-size 25

Rscript tools/run_plant_chemistry_panel.R \
  --mode pilot \
  --plant-csv project_species.csv \
  --out-dir plant_phytochemistry_panel \
  --cache-dir uafR_plant_cache \
  --lotus-index uafR_indexes/LOTUS_lookup_index \
  --sources lotus,npass,knapsack,pubchem,pubmed,pubtator \
  --pilot-count 12
```

Review pilot provider coverage, context extraction, identity resolution,
request counts, cache behavior, and service warnings before running
`--mode discovery`. Exit status 75 is a safe pause. Re-run the same command
later with the same input, output, cache, and index paths.

## Enrichment and similarity gates

Do not send every discovered name directly to rich PubChem/KEGG enrichment.
Review identity resolution, prioritize source-backed structures, and retain a
ledger of deferred or excluded names. Use `estimateTanimotoOutput()` before
requesting all compound-pair rows.

Tanimoto calculations require structure-resolved compounds. Comparable plant
summaries should be filtered by a defensible chemistry scope or group. Keep the
unfiltered summary for audit, but do not treat unlike chemistry as directly
comparable without a documented rationale.

## Minimum handoff

A production handoff should include:

- immutable input and metadata snapshots;
- provider and cache manifests;
- direct, fallback, candidate, and no-hit diagnostics;
- occurrence evidence and review decisions;
- compound identity audit and exclusions;
- enrichment validation;
- comparable-chemistry dictionary and matrices;
- Tanimoto summaries and pairwise-output manifest;
- package version, release artifact, checksums, and session information; and
- explicit statements of unsupported claims.

Run the relevant validation function after copying the final bundle to its
handoff location.
