---
title: "Offline Plant Phytochemistry Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Offline Plant Phytochemistry Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(uafR)
```

## Purpose

This article runs a complete plant-chemistry project without web access. The
included rows are a **simulated teaching fixture**. Their source labels,
citations, occurrence relationships, and project context are synthetic and
must not be treated as evidence about the listed plants.

The fixture exists to demonstrate the package contracts:

- one stable plant universe;
- direct, fallback, candidate, and review-required evidence;
- resolved and unresolved compounds;
- comparable and non-comparable chemistry;
- plant-labeled Tanimoto rows;
- model-ready feature matrices; and
- a validated analysis bundle.

## Load the fixture

```{r load-fixture}
example_dir = system.file(
  "extdata",
  "offline_plant_chemistry",
  package = "uafR"
)
stopifnot(nzchar(example_dir))

plants = read.csv(
  file.path(example_dir, "plant_list.csv"),
  stringsAsFactors = FALSE
)
metadata = read.csv(
  file.path(example_dir, "plant_metadata.csv"),
  stringsAsFactors = FALSE
)
plant_compounds = read.csv(
  file.path(example_dir, "plant_compounds.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
compound_pairs = read.csv(
  file.path(example_dir, "plant_compound_pair_tanimoto.csv"),
  stringsAsFactors = FALSE
)

data.frame(
  plants = nrow(plants),
  metadata_rows = nrow(metadata),
  occurrence_rows = nrow(plant_compounds),
  pair_rows = nrow(compound_pairs)
)
```

Inspect the evidence and identity fields before building outputs:

```{r inspect-input}
plant_compounds[, c(
  "species",
  "compound_name",
  "source_database",
  "evidence_tier",
  "matched_rank",
  "confidence",
  "InChIKey",
  "comparison_scope",
  "comparable_for_matrix",
  "curation_flag"
)]
```

## Build and validate a project

`runPlantChemistryProject()` is the offline/cached project coordinator. Live
discovery is disabled unless `run_discovery = TRUE` is supplied explicitly.

```{r build-project}
project_dir = tempfile("uafr_offline_project_")

project = runPlantChemistryProject(
  plant_list = plants,
  metadata = metadata,
  plant_compounds = plant_compounds,
  plant_compound_pair_tanimoto = compound_pairs,
  output_dir = project_dir,
  project_id = "simulated_offline_example",
  overwrite = TRUE
)

bundle_dir = project$Project$bundle_dir[[1]]
validation = validatePlantChemistryAnalysisBundle(bundle_dir)
stopifnot(identical(
  validation$Summary$ExportReadyStatus[[1]],
  "pass"
))

validation$Summary
```

The validator checks CSV parsing, required columns, artifact relationships,
manifest references, and the bundle’s export-ready status. A passing bundle is
structurally ready for handoff; it does not make the simulated evidence real.

## Inspect evidence and review queues

```{r inspect-evidence}
membership = read.csv(
  file.path(bundle_dir, "03b_PlantCompoundMembershipEnriched.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

grade_summary = read.csv(
  file.path(bundle_dir, "16_EvidenceGradeSummary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

direct = filterPlantEvidenceDirect(membership)
comparable = filterPlantEvidenceComparable(membership)
review_required = filterPlantEvidenceReviewRequired(membership)

grade_summary
data.frame(
  all_rows = nrow(membership),
  direct_structure_resolved = nrow(direct),
  direct_comparable = nrow(comparable),
  review_required = nrow(review_required)
)
```

These filters do not delete the original evidence. They make the analysis rule
explicit while the bundle preserves fallback, candidate, unresolved, and
excluded rows for audit.

## Inspect model-ready features

```{r inspect-features}
feature_manifest = read.csv(
  file.path(bundle_dir, "28_FeatureMatrixManifest.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

feature_manifest[, c(
  "Table",
  "FileName",
  "Mode",
  "RowCount",
  "ColumnCount"
)]

feature_files = file.path(bundle_dir, feature_manifest$FileName)
species_order = lapply(feature_files, function(path) {
  read.csv(
    path,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )$species
})

stopifnot(all(vapply(
  species_order[-1],
  identical,
  logical(1),
  species_order[[1]]
)))
```

Feature matrices are analysis inputs. They are not evidence of mechanism,
efficacy, remediation performance, exposure, pathway activity, or chemistry
measured in a project sample.

## Reproduce outside the vignette

From a repository checkout:

```sh
Rscript tools/build_offline_plant_chemistry_example.R \
  --out-dir offline-plant-example
```

Record `packageVersion("uafR")`, `sessionInfo()`, the input checksums, and the
generated manifest with any downstream analysis.

```{r cleanup, include=FALSE}
unlink(project_dir, recursive = TRUE, force = TRUE)
```
