---
title: "uafR: Reproducible Chemical and Plant-Chemistry Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{uafR: Reproducible Chemical and Plant-Chemistry Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
library(uafR)
```

## Scope

uafR supports three related tasks:

1. preparing tentative GC-MS library-hit tables for downstream analysis;
2. enriching supplied compound identities with source-backed public database
   annotations; and
3. organizing reported plant-compound evidence into auditable summaries,
   comparable-chemistry matrices, structural-similarity summaries, and portable
   project bundles.

These workflows answer different questions. A GC-MS library hit is a tentative
annotation unless confirmed with an appropriate analytical standard. A database
record reports what a source says; it does not establish that a compound occurs
in a new sample. A plant-compound record is evidence of reported occurrence,
not a complete metabolome. Structural similarity is not proof of shared
function, pathway activity, exposure, or efficacy.

Start with the workflow guide:

```{r workflow-guide}
knitr::kable(uafRWorkflowGuide()[, c(
  "user_has", "primary_functions", "stability"
)])
```

## Core GC-MS workflow

`spreadOut()` expects a data frame with the columns `Component.RT`,
`Component.Area`, `Base.Peak.MZ`, `File.Name`, `Compound.Name`, and
`Match.Factor`. It validates those columns and rejects duplicate peak keys
before performing live identity and mass lookups. `mzExacto()` uses the spread
object and a supplied chemical list to aggregate candidate peak areas.
`exactoThese()` can derive a query list from an existing categorate result.

The following is verified against the current function signatures, but is not
run while building this vignette because `spreadOut()` and `mzExacto()` use
live services:

```{r gcms-template, eval=FALSE}
hits = read.csv("gcms_hit_table.csv", stringsAsFactors = FALSE)

spread = spreadOut(hits)
query_compounds = c("methyl salicylate", "octanal")
selected = mzExacto(spread, query_compounds, decontaminate = TRUE)

# A categorate result can also define the query list.
query_from_sources = exactoThese(
  categoratedInput = categorate_result,
  subsetBy = "Database",
  subsetArgs = c("LOTUS", "KEGG")
)
```

Inspect input row counts, duplicate keys, missing values, match factors,
retention times, and area distributions before interpreting `selected`. Record
the instrument, library, search settings, filters, software versions, and
whether authentic standards were used.

## Research-grade compound enrichment

`categorate(detail = "research")` adds PubChem and KEGG profiles, source
coverage, normalized traits, evidence tables, and validation outputs to the
standard categorate tables. Live enrichment should begin with a small smoke
test, a persistent cache, and conservative request spacing:

```{r categorate-live, eval=FALSE}
data("library_data", package = "uafR")

result = categorate(
  compounds = c("aspirin", "caffeine"),
  chemical_library = library_data,
  detail = "research",
  cache = TRUE,
  cache_dir = "uafR_cache",
  throttle = 0.5
)

validation = validateCategorateResult(result)
validation$Summary
result$DataDictionary
```

Do not infer experimental presence from enrichment output. Inspect
`SourceCoverage`, source-specific evidence, unresolved identities, and
validation warnings before selecting analysis variables.

## Complete offline plant-chemistry project

The package includes a small **simulated teaching fixture**. Its plant-compound
rows and source labels are intentionally synthetic and are not evidence that
the listed compounds occur in the listed plants. The fixture exercises the
real project, grading, classification, feature-export, Tanimoto-summary, and
validation code without making network requests.

```{r offline-project}
example_dir = system.file(
  "extdata", "offline_plant_chemistry",
  package = "uafR"
)
stopifnot(nzchar(example_dir))

project_dir = tempfile("uafr_offline_project_")
project = runPlantChemistryProject(
  plant_list = file.path(example_dir, "plant_list.csv"),
  metadata = file.path(example_dir, "plant_metadata.csv"),
  plant_compounds = file.path(example_dir, "plant_compounds.csv"),
  plant_compound_pair_tanimoto = file.path(
    example_dir, "plant_compound_pair_tanimoto.csv"
  ),
  output_dir = project_dir,
  project_id = "simulated_offline_example",
  overwrite = TRUE
)

bundle_dir = project$Project$bundle_dir[[1]]
validation = validatePlantChemistryAnalysisBundle(bundle_dir)
stopifnot(identical(validation$Summary$ExportReadyStatus[[1]], "pass"))
knitr::kable(validation$Summary)
```

The finalized bundle retains the full supplied evidence while making review
and analysis subsets explicit:

```{r inspect-evidence}
membership = read.csv(
  file.path(bundle_dir, "03b_PlantCompoundMembershipEnriched.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
grade_summary = read.csv(
  file.path(bundle_dir, "16_EvidenceGradeSummary.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

direct = filterPlantEvidenceDirect(membership)
comparable = filterPlantEvidenceComparable(membership)
review_required = filterPlantEvidenceReviewRequired(membership)

knitr::kable(grade_summary)
data.frame(
  supplied_rows = nrow(membership),
  direct_structure_resolved_rows = nrow(direct),
  direct_comparable_rows = nrow(comparable),
  review_required_rows = nrow(review_required)
)
```

The filters do not delete evidence from the bundle. They make the selection
rule reproducible. Direct evidence, genus/family fallback, candidate literature
co-mentions, unresolved identities, and review exclusions remain distinct.

## Model-ready feature matrices

The project runner writes count, binary, fraction, and confidence-weighted
feature tables. It retains the same species universe and order across matrix
families, including species with no reported compounds.

```{r inspect-features}
feature_manifest = read.csv(
  file.path(bundle_dir, "28_FeatureMatrixManifest.csv"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
knitr::kable(feature_manifest[, c(
  "Table", "FileName", "Mode", "RowCount", "ColumnCount"
)])

feature_files = file.path(bundle_dir, feature_manifest$FileName)
species_ids = lapply(feature_files, function(path) {
  read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)$species
})
stopifnot(all(vapply(
  species_ids[-1], identical, logical(1), species_ids[[1]]
)))
```

Feature matrices are analysis inputs, not response variables or evidence of
biological efficacy. Preserve coverage, missingness, evidence-grade, and
classification-quality fields when joining them to ecological, phylogenetic,
environmental, or experimental metadata.

## Species-first discovery and local LOTUS use

For real discovery, query direct species records first and keep fallback and
literature candidates separate. Public-source coverage is incomplete and
changes over time. Local indexes and caches are preferred for larger projects.

```{r species-discovery-live, eval=FALSE}
lotus_index = buildLotusIndex(
  input = "path/to/lotus_flat_export.tsv",
  out_file = "uafR_lotus_index/lotus_index.rds"
)

phyto = resolvePlantPhytochemistry(
  plants = c("Salix nigra", "Camellia sinensis"),
  sources = c("lotus", "pubmed", "pubtator"),
  lotus_index = lotus_index,
  taxon_fallback = c("species", "genus"),
  enrich_compounds = TRUE,
  detail = "research",
  cache = TRUE,
  cache_dir = "uafR_plant_cache",
  throttle = 0.5,
  max_pubmed_records = 25
)

validatePlantPhytochemistryResult(phyto)$Summary
phyto$ProviderDiagnostics
phyto$PlantCompoundOccurrences
```

Before a large run, use `planPlantChemistryRun()`, inspect the cache, estimate
pairwise output sizes, and choose batch/cooldown settings. Live-service errors
are not biological no-hits; preserve provider diagnostics and retry queues.

## Structural similarity

uafR calculates PubChem Fingerprint2D Tanimoto similarity only for
structure-resolved compounds. Plant summaries should be filtered by comparable
chemistry scope or group when the scientific question requires like-with-like
comparison. Unknown, unresolved, and non-comparable compounds remain available
for audit but are excluded from comparable summaries by default.

```{r tanimoto-live, eval=FALSE}
estimateTanimotoOutput(
  compound_count = 1000,
  plant_count = 100,
  write_compound_pairs = TRUE
)

similarity = plantChemicalTanimotoSimilarity(
  plant_chemistry = structure_resolved_membership,
  return_compound_pairs = FALSE,
  return_group_compound_pairs = TRUE
)

comparable_summary = plantComparableTanimotoSummary(
  plant_compound_pair_tanimoto = similarity$PlantCompoundTanimoto,
  membership = classified_membership
)
```

Similarity does not establish shared biosynthesis, ecological role, mechanism,
toxicity, or treatment performance. Interpret it with chemical classification,
evidence support, coverage, and the project design.

## Handoff standard

A defensible project handoff should include the input plant list and metadata,
the original and reviewed occurrence evidence, compound identity audit,
provider diagnostics, categorate validation, comparison dictionary, feature
manifest, exported matrices, checksums, methods text, and session information.
Run `validatePlantChemistryAnalysisBundle()` after copying the bundle to its
final location and resolve every failing check before analysis or publication.

```{r cleanup, include=FALSE}
unlink(project_dir, recursive = TRUE, force = TRUE)
```
